---
title: "Accessing ANUClimate v2.0 Data via NCI THREDDS"
author: "Eli R. Bendall, Hawkesbury Institute for the Environment, Western Sydney University"
funding: "This work was supported by Western Sydney University and with funding from the Australian Government under the National Environmental Science Program's Resilient Landscapes Hub. 
orcid: "0000-0002-9120-3457"
date: "`r Sys.Date()`"
note: "Code developed with AI assistance from Claude (Anthropic), https://claude.ai"
output: html_document
---

# Overview

This guide walks through downloading ANUClimate v2.0 monthly climate data
from the NCI THREDDS server for use in the SPEI pipeline on a local machine.
No NCI account is required — the data is publicly accessible via the THREDDS
HTTP file server.

**Data source:** ANUClimate v2.0 (Hutchinson et al., ANU/NCI)
**DOI:** https://doi.org/10.25914/60a10acd183a2
**Coverage:** Australian continent, 0.01° resolution, 1900 to present
**Variables downloaded:** rain, tmax, tmin, srad
**Download period:** January 1980 to December 2024 (required for SPEI pipeline)

---

# Chunk 1 — Install and load packages

```{r setup, message=FALSE, warning=FALSE}
# Install if needed (run once)
# install.packages("here")
# install.packages("lubridate")

library(here)
library(lubridate)
```

---

# Chunk 2 — Define base path and variables

Set the base path to wherever you want to store the downloaded data.
This should be on a drive with at least **50GB free space** — the full
download is approximately 35GB.

```{r paths}
# Set your base download path — adjust to your local drive
base_path <- "/path/to/anuclimate_data"  # Update to your local data directory

# Variables to download
variables <- c("rain", "tmax", "tmin", "srad")

# Date range — 1980 to 2024
# Jan 1980 provides the 30-year reference period (1980-2009)
# Dec 2024 is the end of the output period
dates  <- seq(as.Date("1980-01-01"), as.Date("2024-12-01"), by = "month")
years  <- format(dates, "%Y")
months <- format(dates, "%m")

cat("Date range:", format(dates[1], "%b %Y"), "to", format(tail(dates, 1), "%b %Y"), "\n")
cat("Total months:", length(dates), "\n")
cat("Variables:", paste(variables, collapse = ", "), "\n")
cat("Estimated download size: ~35 GB\n")
```

---

# Chunk 3 — Check available disk space

Check that you have sufficient space before starting the download.

```{r disk-space}
# Check available space on your download drive
# Adjust the path to match your drive
disk_info <- system(paste("df -h", dirname(base_path)), intern = TRUE)
cat(disk_info, sep = "\n")

cat("\nYou need at least 50 GB free (35 GB data + 15 GB working space).\n")
```

---

# Chunk 4 — Create output directories

```{r create-dirs}
# Create one directory per variable
for (var in variables) {
  dir_path <- file.path(base_path, var)
  dir.create(dir_path, recursive = TRUE, showWarnings = FALSE)
  cat("Created:", dir_path, "\n")
}
```

---

# Chunk 5 — Verify server connectivity

Check that the THREDDS server is accessible before starting the full download.

```{r test-connection}
# Test a single file download
test_url  <- "https://thredds.nci.org.au/thredds/fileServer/gh70/ANUClimate/v2-0/stable/month/rain/1980/ANUClimate_v2-0_rain_monthly_198001.nc"
test_file <- file.path(base_path, "rain", "ANUClimate_v2-0_rain_monthly_198001.nc")

if (!file.exists(test_file)) {
  message("Testing server connection...")
  tryCatch({
    download.file(test_url, test_file, mode = "wb", quiet = FALSE)
    cat("Connection OK. Test file downloaded successfully.\n")
    cat("File size:", round(file.size(test_file) / 1e6, 1), "MB\n")
  }, error = function(e) {
    cat("ERROR: Could not connect to THREDDS server.\n")
    cat("Check your internet connection and try again.\n")
    cat("Error message:", e$message, "\n")
  })
} else {
  cat("Test file already exists — server check skipped.\n")
}
```

---

# Chunk 6 — Verify downloaded file structure

Check the test file opened correctly and has the expected structure.

```{r verify-test}
# Requires terra package
if (!requireNamespace("terra", quietly = TRUE)) {
  install.packages("terra")
}
library(terra)

test_file <- file.path(base_path, "rain", "ANUClimate_v2-0_rain_monthly_198001.nc")

if (file.exists(test_file)) {
  r <- rast(test_file)
  cat("=== Test file structure ===\n")
  print(r)
  cat("\nExtent:", as.character(ext(r)), "\n")
  cat("Resolution:", res(r), "\n")
  cat("Variable name:", names(r), "\n")
  cat("Units:", units(r), "\n")
  
  # Check a value at Sydney
  val <- extract(r, data.frame(x = 150.9, y = -33.9))
  cat("Test value at Sydney (Jan 1980):", round(val[1, 2], 2), "mm\n")
} else {
  cat("Test file not found — run Chunk 5 first.\n")
}
```

---

# Chunk 7 — Run the full download

This chunk downloads all files for all four variables from January 1980
to December 2024. **Total size: approximately 35 GB.**

**Expected time:** Several hours depending on your internet connection.
The download is fully restartable — if interrupted, re-run this chunk
and it will skip files that already exist.

```{r download, eval=FALSE}
# Base URL for THREDDS file server
thredds_base <- "https://thredds.nci.org.au/thredds/fileServer/gh70/ANUClimate/v2-0/stable/month"

# Track overall progress
total_files   <- length(variables) * length(dates)
downloaded    <- 0
skipped       <- 0
failed        <- 0
failed_files  <- character(0)

start_time <- Sys.time()

for (var in variables) {
  message("\n--- Downloading: ", var, " ---")

  for (i in seq_along(dates)) {
    yr <- years[i]
    mo <- months[i]

    filename <- paste0("ANUClimate_v2-0_", var, "_monthly_", yr, mo, ".nc")
    url      <- paste0(thredds_base, "/", var, "/", yr, "/", filename)
    destfile <- file.path(base_path, var, filename)

    if (file.exists(destfile)) {
      skipped <- skipped + 1
      next
    }

    tryCatch({
      download.file(url, destfile, mode = "wb", quiet = TRUE)
      downloaded <- downloaded + 1
    }, error = function(e) {
      failed       <<- failed + 1
      failed_files <<- c(failed_files, filename)
      message("  FAILED: ", filename, " — ", e$message)
    })

    # Progress update every 60 files
    total_done <- downloaded + skipped + failed
    if (total_done %% 60 == 0) {
      elapsed  <- round(difftime(Sys.time(), start_time, units = "mins"), 1)
      pct      <- round(total_done / total_files * 100)
      message("  Progress: ", total_done, "/", total_files,
              " (", pct, "%) — ", elapsed, " min elapsed")
    }
  }
}

# Final summary
elapsed_total <- round(difftime(Sys.time(), start_time, units = "mins"), 1)
message("\n=== Download Complete ===")
message("Downloaded: ", downloaded, " files")
message("Skipped (existing): ", skipped, " files")
message("Failed: ", failed, " files")
message("Total time: ", elapsed_total, " minutes")

if (failed > 0) {
  message("\nFailed files:")
  for (f in failed_files) message("  ", f)
  message("\nRe-run this chunk to retry failed files.")
}
```

---

# Chunk 8 — Verify download completeness

After the download finishes, check that all files are present and
have reasonable file sizes.

```{r verify-download}
# Check file counts
cat("=== File count check ===\n")
expected <- length(dates)

for (var in variables) {
  files   <- list.files(file.path(base_path, var), pattern = "\\.nc$")
  n       <- length(files)
  status  <- ifelse(n == expected, "OK", paste("MISSING", expected - n, "files"))
  cat(sprintf("  %-6s: %d / %d files — %s\n", var, n, expected, status))
}

# Check file sizes — should all be ~14-18 MB
cat("\n=== File size check (sample) ===\n")
for (var in variables) {
  files <- list.files(file.path(base_path, var),
                      pattern = "\\.nc$", full.names = TRUE)
  if (length(files) > 0) {
    sizes    <- file.size(files) / 1e6  # MB
    min_size <- round(min(sizes), 1)
    max_size <- round(max(sizes), 1)
    n_small  <- sum(sizes < 5)  # files under 5MB are likely corrupt
    cat(sprintf("  %-6s: %.1f – %.1f MB per file", var, min_size, max_size))
    if (n_small > 0) cat(" — WARNING:", n_small, "files under 5MB (possibly corrupt)")
    cat("\n")
  }
}
```

---

# Chunk 9 — Check all four variables open correctly

```{r verify-variables}
library(terra)

cat("=== Variable verification ===\n")

test_files <- list()
for (var in variables) {
  f <- file.path(base_path, var,
                 paste0("ANUClimate_v2-0_", var, "_monthly_198001.nc"))
  if (file.exists(f)) {
    r              <- rast(f)
    test_files[[var]] <- r
    val            <- extract(r, data.frame(x = 150.9, y = -33.9))
    cat(sprintf("  %-6s: units = %-15s  Sydney Jan 1980 = %.2f\n",
                var, units(r), val[1, 2]))
  } else {
    cat(sprintf("  %-6s: FILE NOT FOUND\n", var))
  }
}

# Confirm all extents match
cat("\n=== Extent consistency check ===\n")
base_ext <- ext(test_files[["rain"]])
for (var in c("tmax", "tmin", "srad")) {
  match <- identical(base_ext, ext(test_files[[var]]))
  cat(sprintf("  rain == %-6s: %s\n", var, ifelse(match, "OK", "MISMATCH — check files")))
}

cat("\nExpected units:\n")
cat("  rain: mm month-1\n")
cat("  tmax: degrees_C\n")
cat("  tmin: degrees_C\n")
cat("  srad: MJ m-2 day-1\n")
```

---

# Chunk 10 — Summary and next steps

```{r summary}
library(terra)

cat("=== Download Summary ===\n")
cat("Base path:", base_path, "\n")
cat("Variables: rain, tmax, tmin, srad\n")
cat("Date range: January 1980 to December 2024\n")
cat("Files per variable:", length(dates), "\n")

# Total size on disk
total_size <- 0
for (var in variables) {
  files      <- list.files(file.path(base_path, var),
                            pattern = "\\.nc$", full.names = TRUE)
  total_size <- total_size + sum(file.size(files))
}
cat("Total size on disk:", round(total_size / 1e9, 1), "GB\n")

cat("\n=== Next Steps ===\n")
cat("1. Open 02_spei_pipeline_local.R\n")
cat("2. Update 'base_path' to match the path above\n")
cat("3. Run the pipeline to generate SPEI-3, 6, 12, and 24 month outputs\n")
cat("\nNote: the SPEI pipeline will create a 'spei' subfolder inside base_path\n")
cat("containing NetCDF and GeoTIFF outputs for the 2010-2024 period.\n")
```

---

# Funding

This work was supported by Western Sydney University and with funding from the Australian Government under the National Environmental Science Program's Resilient Landscapes Hub.

# Data Citation

When using ANUClimate v2.0 data in publications, cite as:

> Hutchinson, M., Xu, T., Kesteven, J., Marang, I., and Evans, B. (2021).
> *ANUClimate v2.0 monthly climate data for the Australian continent,
> 1900 to present*. NCI Australia.
> https://doi.org/10.25914/60a10acd183a2

---

# Variable Descriptions

| Variable | Description | Units |
|---|---|---|
| `rain` | Monthly total rainfall | mm month⁻¹ |
| `tmax` | Monthly mean daily maximum temperature | °C |
| `tmin` | Monthly mean daily minimum temperature | °C |
| `srad` | Monthly mean daily solar radiation | MJ m⁻² day⁻¹ |

---

*This workflow downloads data from the publicly accessible NCI THREDDS server.
No NCI account is required. Data is available for the Australian continent
at 0.01° (~1km) resolution from 1900 to present.*
